\plaintitle

### 2017-02-24 paper/paper.tex
Paper: Several final text improvements

% Paper metadata (use plain text, for PDF inclusion and later
% re-using, if desired).  Use \emtpyauthor when submitting for review
% so you remain anonymous.
\def\plaintitle{Physical Effort and Expressivity in\\ Electronic Musical Instruments}
\def\plainauthor{Tom Gurion}
\def\plainkeywords{effort; expression; electronic music instruments}

@@

Apart from the expressivity of the performer, the ability of the listener to understand the performance might also depend on the relationship between visible effort and the resulted sound \cite{Verdonk2015}.
This idea is discussed thoroughly in the context of live laptop music \cite{Schloss2003}.
In these type of performances, more often than not, the performer stares on the laptop screen, motionless, while operating the music software.
The relationship between cause and effect in these performances lacks any visual cues.
On extreme cases, a listener might question the "liveness" of a performance altogether when unable to tell the difference between live and pre-composed playback.
Besides, visible effort is an important expressive tool in many musical contexts \cite{Schloss2003}.
Consider, for example, a trumpet player and his out-curved veins in the neck while reaching a high note, or a drummer, sweating during intense performance.
These visible expressions usually have significant effect on the audience appreciation of the performance.
Most of the research in the field, however, doesn't confront the theoretical hypotheses using rigorous experiments.
One exceptional experiment based study confirms this hypothesis by showing that when listeners fail to build mental model of how an electronic instrument work they start to rely on facial expression and visible cues to assess errors in performance \cite{Fyans2010}.
Another experiment based study argues against the hypothesis by showing that experienced listeners do able to perceive what the performer is doing from the audio alone \cite{Bown2014}.
Other researchers approach this issue by suggesting the re-education the audience to appreciate ``aural'' performativity instead of visual spectacles \cite{Stuart2003}.
They support their arguments by examples from traditions that accepted the lack of visual cues in musical performances, like tape music concerts, and different variances of sound installations.

Recently, researchers suggested frameworks to assess the expressive possibilities of new interfaces for music performance \cite{Poepel2005}.
In their model, performers communicate emotions by ``coding'' them with a set of musical properties.
Specifically, they mention tempo, sound level, timing, intonation, articulation, timbre, vibrato, tone attacks, tone decays and pauses.
The listeners, on the other hand, ``decode'' these properties to understand the performer expressive intention.
This model for understanding expression in performance was validated in controlled experiment \cite{Gabrielsson1996}.
Now, new interfaces for music performance should provide the means to control these properties to allow the performer to convey emotion.
But controllability is probably not enough.
New interfaces should consider the interface ``efficiency'' - the ratio between the complexity of the resulted sound and the complexity of the controller input \cite{Jorda2004}.
In addition, the ability to master the instrument, questions of controllability versus randomness, and physical feedback should all be taken into account \cite{Dobrian2006}.

@@

Although this controller can be used for different creative purposes, in the context of this study one elastic band controls the pitch and the other controls the amplitude of a sawtooth tone generator.
This type of interaction resembles the design of the theremin \cite{Theremin1996}.
Choosing this type of design provide several advantages for assessing the question of expressivity and physical effort.
First, as noted in \cite{Fyans2010}, it proposes a simple relationship between performer gesture and sound.
Considering that the participants in the study had no former familiarity with the interface, this by itself allow them to understand the interactive behavior good enough to provide valuable feedback.
Second, the added effort required by the Schleikess doesn't complicate the understanding of the instrument but provide the opportunity to test the questions of physical effort and expressivity directly.

@@

Eight participants were invited to play with the Schleikess based synthesizer and discuss their subjective experience.
Apart from the physical interface, figure \ref{fig:pd-interface} shows the interface on the computer screen that was exposed to the participants.
The interface presents three sliders.
The first two control the sensitivity of each elastic band to force.
So, when the sensitivity is set to low value, a major physical effort is needed to raise the pitch or amplitude, but when the sensitivity is set to high value even minor changes affect the output significantly.
The last slider controls the overall volume of the instrument and allows the participant to set it to a comfortable hearing level.

@@

Out of the eight participants in the study, seven understood how the instrument works by trying it out.
I explained the relationship between pulling the elastic bands and the resulted sound to the participant that found it hard to understand.
After the explanation this participant continued with the experiment as usual.

Seven participants explicitly discussed the balance between having range of possibilities and controllability with relationship to the sensitivity of the instrument to force.
This was significantly apparent in the discussion of the sensitivity of the pitch band.
Five participants wanted to maximize the pitch range that the instrument produces by raising the sensitivity, but no participant set the pitch sensitivity to the maximum value.
Four of these participants explicitly explained that they kept the pitch sensitivity below the maximum value to make sure they have precise control of it.
Only one of the five didn't mind having less control over the pitch in favor of larger pitch range.
On the other hand, two participants kept the pitch sensitivity low, and explained this preference by enjoying the lower frequencies notes more.

@@

Three participants mentioned that playing the instrument felt like a physical exercise, is somewhat tedious, or cause them pain over time.
Therefore, two of them preferred to minimize effort or movement while playing.
One participant, attempting to minimize the effort involve in large hand movement, described a completely different playing technique:
Instead of pulling and releasing the bands, the participant held them in constant force and changed the tension by pulling the bands with the little finger.

@@

On the other hand, as well as the required physical effort, the Schleikess also provides physical feedback for the player.
This feedback cannot be separated from the physicality of the interface, and it is an integral part in all of the interfaces of this kind.
I believe that this feedback provided greater controllability and at the same time enhanced the openness of the participants to elements of randomness in their play.
This claims, however, weren't assessed directly by the current study.

>>>>

% Paper metadata (use plain text, for PDF inclusion and later
% re-using, if desired).  Use \emtpyauthor when submitting for review
% so you remain anonymous.
\def\plaintitle{Physical Effort and Expressivity in\\ Electronic Musical Instruments}
\def\plainauthor{Tom Gurion}
\def\plainkeywords{Effort; Expression; Electronic Music Instruments.}

@@

Apart from the expressivity of the performer, the ability of the listener to understand the performance might also depend on the relationship between visible effort and the resulted sound \cite{Verdonk2015}.
This idea is discussed thoroughly in the context of live laptop music \cite{Schloss2003}.
In these type of performances, more often than not, the performer stares on the laptop screen, motionless, while operating music creation software.
The relationship between cause and effect in these performances lacks any visual cues.
On extreme cases, a listener might question the "liveness" of a performance altogether when unable to tell the difference between live and pre-composed playback.
Besides, visible effort is an important expressive tool in many musical contexts \cite{Schloss2003}.
Consider, for example, a trumpet player and his out-curved veins in the neck while reaching a high note, or a drummer, sweating during intense performance.
These visible expressions usually have significant effect on the audience appreciation of the performance.
Most of the research in the field, however, doesn't confront the theoretical hypotheses using rigorous experiments.
One exceptional experiment based study confirms this hypothesis by showing that when listeners fail to build mental model of how an electronic instrument works they start to rely on facial expression and visible cues to assess errors in performance \cite{Fyans2010}.
Another experiment based study argues against the hypothesis by showing that experienced listeners do able to perceive what the performer is doing from the audio alone \cite{Bown2014}.
Other researchers approach this issue by suggesting the re-education of the audience to appreciate ``aural'' performativity instead of visual spectacles \cite{Stuart2003}.
They support their arguments by examples from traditions that accepted the lack of visual cues in musical performances, like tape music concerts, and different variances of sound installations.

Recently, researchers suggested frameworks to assess the expressive possibilities of new interfaces for music performance \cite{Poepel2005}.
They describe expressivity by means of communication between the performer and the audience.
Performers communicate emotions by ``coding'' them with a set of musical properties.
Specifically, they mention tempo, sound level, timing, intonation, articulation, timbre, vibrato, tone attacks, tone decays and pauses as the musical properties that carry the performer expressions.
The listeners, on the other hand, ``decode'' these properties to understand the performer intention.
This model for understanding expression in performance was validated in controlled experiment \cite{Gabrielsson1996}.
Now, new interfaces for music performance should provide the means to control these properties to allow the performer to convey emotion.
But exposing these properties to the performer to control is probably not enough.
New interfaces should consider the interface ``efficiency'' --- the ratio between the complexity of the resulted sound and the complexity of the controller input \cite{Jorda2004}.
In addition, the ability to master the instrument, questions of controllability versus randomness, and physical feedback should all be taken into account \cite{Dobrian2006}.

@@

Although this controller can be used for different creative purposes, in the context of this study one elastic band controls the pitch and the other controls the amplitude of a sawtooth tone generator.
This type of interaction resembles the design of the theremin \cite{Theremin1996}.
Choosing this interface provide several advantages for assessing the question of expressivity and physical effort.
First, as noted in \cite{Fyans2010}, it proposes a simple relationship between performer gesture and sound.
Considering that the participants in the study had no former familiarity with the interface, this simplicity by itself allows them to understand the interactive behavior good enough to provide valuable feedback.
Second, the added effort required by the Schleikess doesn't complicate the understanding of the instrument but provide the opportunity to test the questions of physical effort and expressivity directly.

@@

Eight participants were invited to play with the Schleikess based synthesizer and discuss their subjective experience.
Apart from the physical interface, figure \ref{fig:pd-interface} shows the graphical interface on the computer screen that was exposed to the participants.
The graphical interface presents three sliders.
The first two control the sensitivity of each elastic band to force.
So, when the sensitivity is set to low value, a major physical effort is needed to raise the pitch or amplitude, but when the sensitivity is set to high value even minor changes affect the output significantly.
The last slider controls the overall volume of the instrument and allows the participant to set it to a comfortable hearing level.

@@

Out of the eight participants in the study, seven understood how the instrument works by trying it out.
I explained the relationship between pulling the elastic bands and the resulted sound to the participant that found it hard to understand.
After the explanation the experiment continued as usual.

Seven participants explicitly discussed the balance between having large range of possibilities and controllability with relationship to the sensitivity of the instrument to force.
This was significantly apparent in the discussion on the sensitivity of the pitch band.
Five participants wanted to maximize the pitch range that the instrument produces by raising the sensitivity, but no participant set the pitch sensitivity to the maximum value.
Four of these participants explicitly explained that they kept the pitch sensitivity below the maximum value to make sure they have precise control of it.
Only one of the five didn't mind having less control over the pitch in favor of larger pitch range.
On the other hand, two participants kept the pitch sensitivity low, and explained this preference by enjoying the lower frequencies more.

@@

Three participants mentioned that playing the instrument felt like a physical exercise, is somewhat tedious, or cause them pain over time.
Therefore, two of them preferred to minimize effort or movement while playing.
One participant attempted to minimize the effort involved in large hand movement with unique playing technique:
Instead of pulling and releasing the bands, the participant held them in constant force and changed the tension by pulling the bands with the little finger.

@@

On the other hand, as well as the required physical effort, the Schleikess also provides physical feedback for the player.
This feedback cannot be separated from the physicality of the interface, and it is an integral part in all of the interfaces of this kind --- those that require physical effort.
I believe that this feedback provided greater controllability and at the same time enhanced the openness of the participants to elements of randomness in their play.
This claims, however, weren't assessed directly by the current study.
